# Setting up

## Open RStudio

The first thing we need to do is open RStudio. Do this now. If you currently
have a project open close it by clicking _File > Close project_.

## Naming your package

Before we create our package we need to give it a name. Package names can only
consist of letters, numbers and dots (.) and must start with a letter. While all
of these are allowed it is generally best to stick to just lowercase letters.
Having a mix of lower and upper case letters can be hard for users to remember
(is it **RColorBrewer** or **Rcolorbrewer** or **rcolorbrewer**?). Believe it 
or not choosing a name can be one of the hardest parts of making a package!
There is a balance between choosing a name that is unique enough that it is easy
to find (and doesn't already exist) and choosing something that makes it obvious
what the package does. Acronyms or abbreviations are one option that often works
well. It can be tricky to change the name of a package later so it is worth
spending some time thinking about it before you start.

> **Checking availability**
>
> If there is even a small chance that your package might be used by other
> people it is worth checking that a package with your name doesn't already
> exist. A handy tool for doing this is the **available** package. This package
> will check common package repositories for your name as well as things like
> Urban Dictionary to make sure your name doesn't have some meanings you weren't
> aware of!

At the end of this workshop we want you to have a personal package that you can
continue to add to and use so we suggest choosing a name that is specific to
you. Something like your initials, a nickname or a username would be good
options. For the example code we are going to use `mypkg` and you could use that
for the workshop if you want to.

## Creating your package

To create a template for our package we will use the `usethis::create_package()`
function. All it needs is a path to the directory where we want to create the
package. For the example we put it on the desktop but you should put it
somewhere more sensible.

```{r}
usethis::create_package("~/Desktop/mypkg")
```

You will see some information printed to the console, something like (where USER
is your username):

```{}
✔ Creating 'C:/Users/USER/Desktop/mypkg/'
✔ Setting active project to 'C:/Users/USER/Desktop/mypkg'
✔ Creating 'R/'
✔ Writing 'DESCRIPTION'
Package: mypkg
Title: What the Package Does (One Line, Title Case)
Version: 0.0.0.9000
Authors@R (parsed):
    * First Last <first.last@example.com> [aut, cre] (<https://orcid.org/YOUR-ORCID-ID>)
Description: What the package does (one paragraph).
License: What license it uses
Encoding: UTF-8
LazyData: true
✔ Writing 'NAMESPACE'
✔ Writing 'mypkg.Rproj'
✔ Adding '.Rproj.user' to '.gitignore'
✔ Adding '^mypkg\\.Rproj$', '^\\.Rproj\\.user$' to '.Rbuildignore'
✔ Opening 'C:/Users/USER/Desktop/mypkg/' in new RStudio session
✔ Setting active project to '<no active project>'
```

You will see something similar whenever we run a **usethis** command. Green
ticks indicate that a step has been completed correctly. If you ever see a red
dot that means that there is something **usethis** can't do for you and you will
need to follow some instructions to do it manually. At the end a new RStudio
window with your package should open. In this window you should see the
following files:

* `DESCRIPTION` - The metadata file for your package. We will fill this in next
  and it will be updated as we develop our package.
* `NAMESPACE` - This file describes the functions in our package. Traditionally
  this has been a tricky file to get right but the modern development tools
  mean that we shouldn't need to edit it manually. If you open it you will see
  a message telling you not to.
* `R/` - This is the directory that will hold all our R code.

These files are the minimal amount that is required for a package but we will
create other files as we go along. Some other useful files have also been
created by **usethis**.

* `.gitignore` - This is useful if you use git for version control.
* `.Rbuildignore` - This file is used to mark files that are in the directory
  but aren't really part of the package and shouldn't be included when we build
  it. Most of the time you won't need to worry about this as **usethis** will
  edit it for you.
* `mypkg.Rproj` - The RStudio project file. Again you don't need to worry about
  this.

## Filling in the DESCRIPTION

The `DESCRIPTION` file is one of the most important parts of a package. It
contains all the metadata about the package, things like what the package is
called, what version it is, a description, who the authors are, what other
packages it depends on etc. Open the `DESCRIPTION` file and you should see
something like this (with your package name).

```{}
Package: mypkg
Title: What the Package Does (One Line, Title Case)
Version: 0.0.0.9000
Authors@R: 
    person(given = "First",
           family = "Last",
           role = c("aut", "cre"),
           email = "first.last@example.com",
           comment = c(ORCID = "YOUR-ORCID-ID"))
Description: What the package does (one paragraph).
License: What license it uses
Encoding: UTF-8
LazyData: true
```

### Title and description

The package name is already set correctly but most of the other fields need to
be updated. First let's update the title and description. The title should be
a single line in Title Case that explains what your package is. The description
is a paragraph which goes into a bit more detail. For example you could write
something like this:

```{}
Package: mypkg
Title: My Personal Package
Version: 0.0.0.9000
Authors@R: 
    person(given = "First",
           family = "Last",
           role = c("aut", "cre"),
           email = "first.last@example.com",
           comment = c(ORCID = "YOUR-ORCID-ID"))
Description: This is my personal package. It contains some handy functions that
    I find useful for my projects.
License: What license it uses
Encoding: UTF-8
LazyData: true
```

### Authors

The next thing we will update is the Authors@R field. There are a couple of ways
to define the author for a package but Authors@R is the most flexible. The
example shows us how to define an author. You can see that the example person
has been assigned the author ("aut") and creator ("cre") roles. There must be
at least one author and one creator for every package (they can be the same
person) and the creator must have an email address. There are many possible
roles (including woodcutter ("wdc") and lyricist ("lyr")) but the most important
ones are:

* cre: the creator or maintainer of the package, the person who should be
  contacted with there are problems
* aut: authors, people who have made significant contributions to the package
* ctb: contributors, people who have made smaller contributions
* cph: copyright holder, useful if this is someone other than the creator (such
  as their employer)
  
> **Adding an ORCID**
>
> If you have an ORCID you can add it as a comment as shown in the example.
> Although not an official field this is recognised in various places (including
> CRAN) and is recommended if you want to get academic credit for your package
> (or have a common name that could be confused with other package authors).

Update the author information with your details. If you need to add another
author simply concatenate them using `c()` like you would with a normal vector.

```{}
Package: mypkg
Title: My Personal Package
Version: 0.0.0.9000
Authors@R: c(
    person(given = "Package",
           family = "Creator",
           role = c("aut", "cre"),
           email = "package.creator@mypkg.com"),
    person(given = "Package",
           family = "Contributor",
           role = c("ctb"),
           email = "package.contributor@mypkg.com")
    )
Description: This is my personal package. It contains some handy functions that
    I find useful for my projects.
License: What license it uses
Encoding: UTF-8
LazyData: true
```

### License

The last thing we will update now is the software license. The describes how
our code can be used and without one people must assume that it can't be used
at all! It is good to be as open and free as you can with your license to make
sure your code is as useful to the community as possible. For this example we
will use the MIT license which basically says the code can be used for any
purpose and doesn't come with any warranties. There are templates for some of
the most common licenses included in **usethis**.

```{r}
usethis::use_mit_license("Your Name")
```

This will update the license field.

```{}
Package: mypkg
Title: My Personal Package
Version: 0.0.0.9000
Authors@R: c(
    person(given = "Package",
           family = "Creator",
           role = c("aut", "cre"),
           email = "package.creator@mypkg.com"),
    person(given = "Package",
           family = "Contributor",
           role = c("ctb"),
           email = "package.contributor@mypkg.com")
    )
Description: This is my personal package. It contains some handy functions that
    I find useful for my projects.
License: MIT + file LICENSE
Encoding: UTF-8
LazyData: true
```

It will also also create two new files, `LICENSE.md` which contains the text
of the MIT license (it's very short if you want to give it a read) and `LICENSE`
which simply contains:

```{}
YEAR: 2019
COPYRIGHT HOLDER: Your Name
```

There are various other licenses you can use but make sure you choose one
designed for software not other kinds of content. For example the Creative
Commons licenses are great for writing or images but aren't designed for code.
For more information about different licenses and what they cover have a look at
http://choosealicense.com/ or https://tldrlegal.com/. For a good discussion
about why it is important to declare a license read this blog post by Jeff
Attwood http://blog.codinghorror.com/pick-a-license-any-license/. 
