# Functions

## Adding a function

Now that our package is all set up it's time to add our first function! We can
use the `usethis::use_r()` function to set up the file. Our function is going
to be about colours so we will use that as the name of the R file.

```{r}
usethis::use_r("colours")
```

> **Organising your code**
>
> There are no rules about how to organise your functions into different files
> but you want generally want to group similar functions into a file with a
> a clear name. Having all of your functions in a single file isn't great, but
> neither is having a separate file for each function. A good rule of thumb is
> that if you are finding it hard to locate a function you might need to move
> it to a new file. There are two shortcuts for finding functions in RStudio,
> selecting a function name and pressing **F2** or pressing **Ctrl + .** and
> searching for the function.

As an example we are going to write a function that takes the red, green and
blue values for a colour and returns a given number of shades. Copy the
following code into your R file and save it (you can ignore the comments if you
want to, they are just there to explain how the function works).

```{r}
make_shades <- function(colour, n, lighter = TRUE) {
    # Convert the colour to RGB
    colour_rgb <- grDevices::col2rgb(colour)[, 1]
    
    # Decide if we are heading towards white or black
    if (lighter) {
        end <- 255
    } else {
        end <- 0
    }

    # Calculate the red, green and blue for the shades
    # we calculate one extra point to avoid pure white/black
    red <- seq(colour_rgb[1], end, length.out = n + 1)[1:n]
    green <- seq(colour_rgb[2], end, length.out = n + 1)[1:n]
    blue <- seq(colour_rgb[3], end, length.out = n + 1)[1:n]
    
    # Convert the RGB values to hex codes
    shades <- grDevices::rgb(red, green, blue, maxColorValue = 255)
    
    return(shades)
}
```

## Using the function

Now that we have a function we want to see if it works. Usually when we write
a new function we load it by copying the code to the console or sourcing the
R file. When we are developing a package we want to try and keep our 
environment empty so that we can be sure we are only working with objects inside
the package. Instead we can load functions using `devtools::load_all()`.

```{r}
devtools::load_all()
```

The function doesn't appear in the environment, just like all the functions in
a package don't appear in the environment when we load it using `library()`.
But if we try to use it the function should work.

```{r}
make_shades("goldenrod", 5)
```

Congratulations, you now have a functional package! In the next section we
will perform some checks to see if we have forgotten anything.
