# Other documentation

In a previous section we documented our functions using Roxygen comments but
there are a few other kinds of documentation we should have.

## Package help file

Users can find out about our functions using `?function-name` but what if they
want to find out about the package itself? There is some information in the
`DESCRIPTION` but that can be hard to access. Let's add a help file for the
pacakge.

```{r}
usethis::use_package_doc()
```

```{}
✔ Writing 'R/mypkg-package.R'
```

This creates a special R file for us called `mypkg-package.R`. The contents of
this file doesn't look like much it is understood by **devtools** and
**roxygen2**.

```{r}
#' @keywords internal
"_PACKAGE"

# The following block is used by usethis to automatically manage
# roxygen namespace tags. Modify with care!
## usethis namespace: start
## usethis namespace: end
NULL
```

Run `devtools::document()`. 

```{r}
devtools::document()
```

```{}
Updating mypkg documentation
Writing NAMESPACE
Loading mypkg
Writing NAMESPACE
Writing mypkg-package.Rd
```

We can see that a new `.Rd` file has been created and we can view the contents
using `?mypkg`. The information here has been automatically pulled from the
`DESCRIPTION` file so we only need to update it in one place.

## Vignettes

The documentation we have written so far explains how individual functions
work in detail but it doesn't show what the package does as a whole. Vignettes
are short tutorials that explain what the package is designed for and how 
different functions can be used together. There are different ways to write
vignettes but usually they are R Markdown files. We can create a vignette with
`usethis::use_vignette()`. There can be multiple vignettes but it is common
practice to start with one that introduces the whole package.

> **What is R Markdown?**
>
> Markdown is a simple markup language that makes it possible to write documents
> with minimal formatting. See _Help_ > _Markdown Quick Reference_ in RStudio
> for a quick guide to how this formatting works. R Markdown adds chunks of R
> code that are run and the output included in the final document. 

```{r}
usethis::use_vignette("mypkg")
```

```{}
✔ Adding 'knitr' to Suggests field in DESCRIPTION
✔ Setting VignetteBuilder field in DESCRIPTION to 'knitr'
✔ Adding 'inst/doc' to '.gitignore'
✔ Creating 'vignettes/'
✔ Adding '*.html', '*.R' to 'vignettes/.gitignore'
✔ Adding 'rmarkdown' to Suggests field in DESCRIPTION
✔ Writing 'vignettes/mypkg.Rmd'
● Modify 'vignettes/mypkg.Rmd'
```

Because this is our first vignette **usethis** has added some information to
the `DESCRIPTION` file including adding the **knitr** package as a suggested
dependency. It also creates a `vignettes/` directory and opens our new
`mypkg.Rmd` file.

````
---
title: "mypkg"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mypkg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE} `r ''`
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup} `r ''`
library(mypkg)
```

````

If you are familiar with R Markdown you might note some unusual content in the
header. This is important for the vignette to build properly. There are also
some **knitr** options set which are the convention for vignettes.

Let's add a short example of how to use our package.

````
---
title: "mypkg"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mypkg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE} `r ''`
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup} `r ''`
library(mypkg)
```

# Introduction

This is my personal package. It contains some handy functions that I find useful
for my projects.

# Colours

Sometimes you want to generate shades of a colour. The `make_shades()` function
makes this easy!

```{r} `r ''`
shades <- make_shades("goldenrod", 5)
```

If you want to see what the shades look like you can plot them using
`plot_colours()`.

```{r} `r ''`
plot_colours(shades)
```

This function is also useful for viewing any other palettes.

```{r} `r ''`
plot_colours(rainbow(5))
```
````

To see what the vignette looks like run `devtools::build_vignettes()`. Asking
**devtools** to build the vignette rather than rendering it in another way
(such as the _Knit_ button in RStudio) makes sure that we are using the
development version of the package rather than any version that is installed.

```{r}
devtools::build_vignettes()
```

```{}
Building mypkg vignettes
--- re-building 'mypkg.Rmd' using rmarkdown
--- finished re-building 'mypkg.Rmd'

Moving mypkg.html, mypkg.R to doc/
Copying mypkg.Rmd to doc/
Building vignette index
```

This creates a new directory called `doc/` that contains the rendered vignette.
Click on the `mypkg.html` file and open it in your browser.

If you want to use any other packages in your vignette that the package doesn't
already depend on you need to add them as a suggested dependency.

## README

If you plan on sharing the source code rather than the built package it is
useful to have a README file to explain what the package is, how to install and
use it, how to contribute etc. We can create a template with
`usethis::use_readme_md()` (if we wanted to and R Markdown file with R code and
output we might use `usethis::use_readme_md()` instead).

```{r}
usethis::use_readme_md()
```

```{}
✔ Writing 'README.md'
● Modify 'README.md'
```

````
# mypkg

<!-- badges: start -->
<!-- badges: end -->

The goal of mypkg is to ...

## Installation

You can install the released version of mypkg from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("mypkg")
```

## Example

This is a basic example which shows you how to solve a common problem:

``` r
library(mypkg)
## basic example code
```
````

There are the comments near the top that mention badges and you might have seen
badges (or shields) on README files in code repositories before. There are
several **usethis** functions for adding badges. For example we can mark this
package as been at the experimental stage using
`usethis::use_lifecycle_badge()`.

```{r}
usethis::use_lifecycle_badge("experimental")
```

```
# mypkg

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

The goal of mypkg is to ...
```

The rest of the template isn't very useful so replace it with something better.

## Package website

If you have a publicly available package it can be useful to have a website
displaying the package documentation. It gives your users somewhere to go and
helps your package appear in search results. Luckily this is easily achieved
using the **pkgdown** package. If you have it installed you can set it up with
**usethis**.

```{r}
usethis::use_pkgdown()
```

