# Versioning

We now have at least something for all of the major parts of our package.
Whenever you reach a milestone like this it is good to update the package
version. Having a good versioning system is important when it comes to things
like solving user issues. Version information is recorded in the `DESCRIPTION`
file. This is what we have at the moment.

```{}
Version: 0.0.0.9000
```

This version number follows the format `major.minor.patch.dev`. The different
parts of the version represent different things:

* `major` - A significant change to the package that would be expected to break
  users code. This is updated very rarely when the package has been redesigned
  in some way.
* `minor` - A minor version update means that new functionality has been added
  to the package. It might be new functions to improvements to existing
  functions that are compatible with most existing code.
* `patch` - Patch updates are bug fixes. They solve existing issues but don't
  do anything new.
* `dev` - Dev versions are used during development and this part is missing from
  release versions. For example you might use a dev version when you  give
  someone a beta version to test. A package with a dev version can be expected
  to change rapidly or have undiscovered issues.

Now that we know how this system works let's increase our package version.

```{r}
usethis::use_version()
```

```{}
Current version is 0.0.0.9000.
Which part to increment? (0 to exit) 

1: major --> 1.0.0
2: minor --> 0.1.0
3: patch --> 0.0.1
4:   dev --> 0.0.0.9001

Selection:
```

The prompt asks us which part of the version we want to increment. We have added
some new functions so let's make a new minor version.

```{}
Selection: 2
✔ Setting Version field in DESCRIPTION to '0.1.0'
```

Whenever we update the package version we should record what changes have been
made. We do this is a `NEWS.md` file.

```{r}
usethis::use_news_md()
```

```{}
✔ Writing 'NEWS.md'
● Modify 'NEWS.md'
```

Modify the file to record what we have done during the workshop.

```{}
# mypkg 0.1.0

* Created the package
* Added the `make_shades()` function
* Added the `plot_colours()` function
* Added a vignette
```
