# Building, installing and releasing

If you want to start using your package in other projects the simplest thing
to do is run `devtools::install()`. This will install your package in the same
way as any other package so that it can be loaded with `library()`. However this
will only work on the computer you are developing the package on. If you want
to share the package with other people (or other computers you work on) there
are a few different options.

## Building

One way to share your package is to manually transfer it to somewhere else. But
rather then copying the development directory what you should share is a
prebuilt package archive. Running `devtools::build()` will bundle up your
package into a `.tar.gz` file without any of the extra bits required during
development. This archive can then be transferred to wherever you need it and
installed using `install.packages("mypkg.tar.gz", repos = NULL)` or `R CMD
INSTALL mypkg.tar.gz` on the command line. While this is fine if you just want
to share the package with yourself or a few people you know, it doesn't work if
you want it to be available to the general community.

## Official repositories

### CRAN

The most common repository for public R packages is the Comprehensive R Archive
Network (CRAN). This is where packages are usually downloaded from when you
use `install.packages()`. Compared to similar repositories for other programming
languages getting your package accepted to CRAN means meeting a series of
requirements. While this makes the submission process more difficult it gives
users confidence that your package is reliable and will work on multiple
platforms. It also makes your package much easier to install for most users and
makes it more discoverable. The details of the CRAN submission process are
beyond the scope of this workshop but it is very well covered in the "Release"
section of Hadley Wickham's "R packages" book
(http://r-pkgs.had.co.nz/release.html) and the CRAN section of Karl Broman's
"R package primer" (https://kbroman.org/pkg_primer/pages/cran.html). You should
also read the offical CRAN submission checklist 
https://cran.r-project.org/web/packages/submission_checklist.html. The CRAN
submission process has a reputation for being prickly and frustrating to go
through but it is important to remember that the maintainers are volunteering
their time to do this for thousands of packages. Because of their hard work
CRAN is a large part of why R is so successful.

### Bioconductor

If your package is designed for analysing biological data you might want to
submit it to Bioconductor rather than CRAN. While Bioconductor has a smaller
audience it is more specialised and is often the first place researchers in the
life sciences look. Building a Bioconductor package also means that you can
take advantage of the extensive ecosystem of existing objects and packages for
handling biological data types. While there are lots of advantages to having
your package on Bioconductor the coding style is slightly different to what is
often used for CRAN packages. If you think you might want to submit your
package to Bioconductor in the future have a look at the Bioconductor package
guideline (https://www.bioconductor.org/developers/package-guidelines/) and the
how to guide to building a Bioconductor package
(https://www.bioconductor.org/developers/how-to/buildingPackagesForBioc/). The
Bioconductor submission process is conducted through GitHub 
(https://bioconductor.org/developers/package-submission/). The Bioconductor
maintainers will guide you through the process and make suggestions about how
to improve your package and integrate it with other Bioconductor packages.
Unlike CRAN which uploads packages all year round Bioconductor has two annual
releases, usually in April and October. This means that all the packages in a
release are guaranteed to be compatible with each other but make sure you
submit in time or you will have to wait another six months for your package to
be available to most users.

### rOpenSci

rOpenSci is not a package repository as such but an organisation that reviews
and improves R packages. Packages that have been accepted by rOpenSci should
meet a certain set of standards. By submitting your package to rOpenSci you will
get it reviewed by experienced programmers who can offer suggestions on how to
improve it. If you are accepted you will receive assistance with maintaining
your package and it will be promoted by the organisation. Have a look at their
submission page for more details https://github.com/ropensci/software-review.

## Code sharing websites

Uploading your package to a code sharing website such as GitHub, Bitbucket or
GitLab offers a good compromise between making your package available and
going through an official submission process. This is a particularly good
option for packages that are still in early development and are not ready to
be submitted to one of the major repositories. Making your package available on
one of these sites also gives it a central location for people to ask questions
and submit issues. Code sharing websites are usually accessed through the git
version control system. If you are unfamiliar with using git on the command line
there are functions in **usethis** that can run the commands for you from R. The
following steps will take you through uploading a package to GitHub but they are
similar for other websites. If you don't already have a GitHub account create
one here https://github.com/join.

### Set up git

First we need to configure our git details.

```{r}
usethis::use_git_config(user.name = "Your Name", user.email = "your@email.com")
```

The email address should be the same one you used to sign up to GitHub. Now we
can set up git in our package repository.

```{r}
usethis::use_git()
```

```{}
✔ Initialising Git repo
There are 13 uncommitted files:
* '.gitignore'
* '.Rbuildignore'
* 'DESCRIPTION'
* 'LICENSE'
* 'LICENSE.md'
* 'man/'
* 'mypkg.Rproj'
* 'NAMESPACE'
* 'NEWS.md'
* 'R/'
* 'README.md'
* 'tests/'
* 'vignettes/'
Is it ok to commit them?

1: Not now
2: Yup
3: Negative

Selection: 2
✔ Adding files
✔ Commit with message 'Initial commit'
● A restart of RStudio is required to activate the Git pane
Restart now?

1: No
2: No way
3: Yes

Selection: 1
```

If you are already familiar with git this should make sense to you. If not, what
this step does (in summary) is set up git and save the current state of the
package. If you chose to restart RStudio you will see a new git pane that can
be used to complete most of the following steps by pointing and clicking.

### Connect to GitHub

The next step is to link the directory on your computer with a repository on
GitHub. First we need to create a special access token. The following command
will open a GitHub website.

```{r}
usethis::use_github()
```

```{}
✔ Opening URL 'https://github.com/settings/tokens/new?scopes=repo,gist&description=R:GITHUB_PAT'
● Call `usethis::edit_r_environ()` to open '.Renviron'.
● Store your PAT with a line like:
  GITHUB_PAT=xxxyyyzzz
  [Copied to clipboard]
● Make sure '.Renviron' ends with a newline!
```

Click the "Generate token" button on the webpage and then copy the code on the
next page. As it says you can only view this once so be careful to copy it now
and don't close the page until you are finished. When you have the code follow
the rest of the instructions from **usethis**.

```{r}
usethis::edit_r_environ()
```

```{}
● Modify 'C:/Users/Luke/Documents/.Renviron'
● Restart R for changes to take effect
```

Edit the file to look something like this (with your code).

```{}
GITHUB_PAT=YOUR_CODE_GOES_HERE
```

Save it then restart R by clicking the _Session_ menu and selecting _Restart R_
(or using **Ctrl+Shift+F10**).

```{}
Restarting R session...
```

Copying that code and adding it to your `.Renviron` gives R on the computer you
are using access to your GitHub repositories. If you move to a new computer you
will need to do this again. Now that we have access to GitHub we can create a
repository for our packages.

```{r}
usethis::use_github()
```

```{}
✔ Setting active project to 'C:/Users/Luke/Desktop/mypkg'
✔ Checking that current branch is 'master'
Which git protocol to use? (enter 0 to exit) 

1: ssh   <-- presumes that you have set up ssh keys
2: https <-- choose this if you don't have ssh keys (or don't know if you do)

Selection: 2
● Tip: To suppress this menu in future, put
  `options(usethis.protocol = "https")`
  in your script or in a user- or project-level startup file, '.Rprofile'.
  Call `usethis::edit_r_profile()` to open it for editing.
● Check title and description
  Name:        mypkg
  Description: My Personal Package
Are title and description ok?

1: For sure
2: No way
3: Negative

Selection: 1
✔ Creating GitHub repository
✔ Setting remote 'origin' to 'https://github.com/lazappi/mypkg.git'
✔ Adding GitHub links to DESCRIPTION
✔ Setting URL field in DESCRIPTION to 'https://github.com/lazappi/mypkg'
✔ Setting BugReports field in DESCRIPTION to 'https://github.com/lazappi/mypkg/issues'
✔ Pushing 'master' branch to GitHub and setting remote tracking branch
✔ Opening URL 'https://github.com/lazappi/mypkg'
```

Respond to the prompts from **usethis** about the method for connecting to
GitHub and the title and description for the repository. When everthing is done
a website should open with your new package repository. Another thing this
function does is add some extra information to the description that let's people
know where to find your new website.

```r}
URL: https://github.com/user/mypkg
BugReports: https://github.com/user/mypkg/issues
```

### Installing from GitHub

Now that your package is on the internet anyone can install it using the
`install_github()` function in the **remotes** package (which you should
already have installed as a dependency of **devtools**). All you need to give
it is the name of the user and repository.

```{r}
remotes::install_github("user/mypkg")
```

If you are familiar with git you can install from a particular branch, tag or
commit by adding that after `@`.

```{r}
remotes::install_github("user/mypkg@branch_name")
remotes::install_github("user/mypkg@tag_id")
remotes::install_github("user/mypkg@commit_sha")
```

### Updating GitHub

After you make improvements to your package you will probably want to update the
version that is online. To do this you need to learn a bit more about git. Jenny
Bryan's "Happy Git with R" tutorial (https://happygitwithr.com) is a great place
to get started but the (very) quick steps in RStudio are:

1. Open the _Git_ pane (it will be with _Environment_, _History_ etc.)
2. Click the check box next to each of the listed files
3. Click the _Commit_ button
4. Enter a message in the window that opens and click the _Commit_ button
5. Click the _Push_ button with the green up arrow

Refresh the GitHub repository website and you should see the changes you have
made.
